### 2025-01-31 paper/preprint/paper.tex
uff this is going too slow

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Explanations, Explainable AI},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
van Deursen}\\Faculty of Electrical Engineering, Mathematics and
Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual Explanations (CE) have emerged as a popular method to
explain the predictions made by opaque machine learning models in a
post-hoc fashion. We propose a novel approach that leverages
counterfactuals during the training phase of models.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep 
Explainable AI

@@

@@

(Wachter, Mittelstadt, and Russell 2017; Joshi et al. 2019; Altmeyer et
al. 2024)

@@

Sauer and Geiger (2021) generate counterfactual images for MNIST and
ImageNet through independent mechanisms (IM): each IM learns
class-conditional input distributions over a specific lower-dimensional,
semantically meaningful factor, such as \emph{texture}, \emph{shape} and
\emph{background}. The demonstrate that using these generated
counterfactuals during classifier training improves model robustness.
Similarly, Abbasnejad et al. (2020) argue that counterfactuals represent
potentially useful training data in machine learning, especially in
supervised settings where inputs may be reasonably mapped to multiple
outputs. They, too, demonstrate the augmenting the training data of
image classifiers can improve generalization.

Teney, Abbasnedjad, and Hengel (2020) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on, firstly, identifying
similar input samples with different annotations and, secondly, ensuring
that the gradient of the classifier aligns with the vector between pairs
of counterfactual inputs using the cosine distance as a loss function
(referred to as \emph{gradient supervision}) (\textbf{\emph{this might
be useful for our task as well}}). In the natural language processing
(NLP) domain, counterfactuals have similarly been used to improve models
through data augmentation: Wu et al. (2021), propose POLYJUICE, a
general-purpose counterfactual generator for language models. They
demonstrate empirically that augmenting training data through POLYJUICE
counterfactuals improves robustness in a number of NLP tasks.

@@

Freiesleben (2022) propose two definitional differences between
Adversarial Examples (AE) and Counterfactual Explanations (CE): firstly,
and more importantly according to the authors, the term AE implies
missclassification, which is not the case for CE (\textbf{\emph{this
might be a useful notion for use to distinguish between adversarials and
explanations during training}}); secondly, they argue that closeness
plays a more critical role in the context of CE but confess that even
counterfactuals that are not close might be relevant explanations.
Pawelczyk et al. (2022) show that CE and AE are equivalent under certain
conditions and derive upper bounds on the distances between them.

@@

Guo, Nguyen, and Yadav (2023) are the first to propose end-to-end
training pipeline that includes counterfactual explanations as part of
the training prodeduce. In particular, they propose a specific network
architecture that includes a predictor and CE generator network
(\textbf{\emph{akin a GAN?}}), where the parameters of the CE generator
network are learnable. Counterfactuals are generated during each
training iteration and fed back to the predictor network
(\textbf{\emph{here we are aligned}}). In contrast, we impose no
restrictions on the neural network architecture at all.
(\textbf{\emph{to ensure the one-hot encoding of categorical features is
maintained, they simple use softmax (might be interesting for CE.jl)}})
Interestingly, the authors find that their approach is sensitive to the
choice of the loss function: only MSE seems to lead to good performance.
They also demonstrate theoretically, that the objective function is
difficult to optimize due to divergent gradients and suffers from poor
adversarial robustness. (\textbf{\emph{because partial gradients with
respect to the classification loss component and the counterfactual
validity component point in opposite directions}}). To mitigate these
issues, the authors use block-wise gradient descent: they first update
with respect to classification loss and then use a second update with
respect to the other loss components (\textbf{\emph{this might be useful
for our task as well}}). Ross, Lakkaraju, and Bastani (2024) propose a
way to train models that are guaranteed to provide recourse for
individuals with high probability. The approach builds on adversarial
training (\textbf{\emph{here we are aligned}}), where in this context
adversarial examples are actively encouraged to exist, but only target
attacks with respect to the positive class. The proposed method allows
for imposing a set of actionable recourse ex-ante: for example, users
can impose mutability constraints for features (\textbf{\emph{here we
are aligned}}). (\textbf{\emph{To solve their objective function more
efficiently, they use a first-order Taylor approximation to approximate
the recourse loss component (might be applicable in our case)}})

Luu and Inoue (2023) introduce Counterfactual Adversarial Training (CAT)
with intention of improving generalization and robustness of language
models. Specifically, they propose to proceed as follows: firstly,
identify training samples that are subject to high predictive
uncertainty (entropy); secondly, generate counterfactual explanations
for those samples; and, finally, finetune the model on the augmented
dataset that includes the generated counterfactuals.

@@

@@

@@

@@

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Counterfactual Training: Teaching Models Plausible and Actionable Explanations},
  pdfauthor={Patrick Altmeyer; Arie van Deursen; Cynthia C. S. Liem},
  pdfkeywords={Counterfactual Explanations, Explainable
AI, Representation Learning},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\usepackage{lineno}
\linenumbers
\newcommand{\runninghead}{A Preprint }
\renewcommand{\runninghead}{Counterfactual Training (A Preprint) }
\title{Counterfactual Training: Teaching Models Plausible and Actionable
Explanations}
\def\asep{\\\\\\ } % default: all authors on same column
\def\asep{\And }
\author{\textbf{Patrick
Altmeyer}~\orcidlink{0000-0003-4726-8613}\\Faculty of Electrical
Engineering, Mathematics and Computer Science\\Delft University of
Technology\\\\\href{mailto:p.altmeyer@tudelft.nl}{p.altmeyer@tudelft.nl}\asep\textbf{Arie
van Deursen}\\Faculty of Electrical Engineering, Mathematics and
Computer Science\\Delft University of Technology\\\\\asep\textbf{Cynthia
C. S. Liem}\\Faculty of Electrical Engineering, Mathematics and Computer
Science\\Delft University of Technology\\\\}
\date{}
\begin{document}
\maketitle
\begin{abstract}
Counterfactual Explanations (CE) have emerged as a popular tool to
explain predictions made by opaque machine learning models: they explain
how factual inputs need to change in order for some fitted model to
produce some desired output. Much existing research has focused on
identifying explanations that are not only valid but also deemed
desirable with respect to the underlying data and stakeholder
requirements. Recent work has shown that under this premise, the task of
learning desirable explanations is effectively reassigned from the model
itself to the (post-hoc) counterfactual explainer. Building on that
work, we propose a novel model objective that leverages counterfactuals
during the training phase (ad-hoc) in order to minimize the divergence
between learned representations and desirable explanations. Through
extensive experiments, we demonstrate that our proposed methodology
facilitates training models that inherently deliver desirable
explanations while maintaining high predictive performance.
\end{abstract}
{\bfseries \emph Keywords}
\def\sep{\textbullet\ }
Counterfactual Explanations \sep Explainable AI \sep 
Representation Learning

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern AIs are tasked with learning these representations from scratch,
guided by narrow objectives such as predictive accuracy
(\citeproc{ref-goodfellow2016deep}{I. Goodfellow, Bengio, and Courville
2016}). Modern advances in computing have made it possible to provide
such AIs with ever greater degrees of freedom to achieve that task,
which has often led them to outperform traditionally more parsimonious
models. Unfortunately, in doing so they have also learned increasingly
complex representations that we can no longer easily interpret.

This trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, I. J. Goodfellow, Shlens, and Szegedy
(\citeproc{ref-goodfellow2014explaining}{2014}) showed that artificial
neural networks (ANN) are highly sensitive to adversarial examples (AE):
counterfactuals of model inputs that yield vastly different model
predictions despite being semantically indifferent from their factual
counterparts. Despite continued research efforts towards making models
more robust to such counterfactuals, the issue remains unresolved even
for models that are considered to small by today's standards
(\citeproc{ref-kolter2023keynote}{Kolter 2023}). It seems that robust
representation learning is fundamentally at odds with large degrees of
freedom.

Wilson (\citeproc{ref-wilson2020case}{2020})

Rudin (\citeproc{ref-rudin2019stop}{2019})

@@

(\citeproc{ref-wachter2017counterfactual}{Wachter, Mittelstadt, and
Russell 2017}; \citeproc{ref-joshi2019realistic}{Joshi et al. 2019};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024})

@@

Sauer and Geiger (\citeproc{ref-sauer2021counterfactual}{2021}) generate
counterfactual images for MNIST and ImageNet through independent
mechanisms (IM): each IM learns class-conditional input distributions
over a specific lower-dimensional, semantically meaningful factor, such
as \emph{texture}, \emph{shape} and \emph{background}. The demonstrate
that using these generated counterfactuals during classifier training
improves model robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization.

Teney, Abbasnedjad, and Hengel (\citeproc{ref-teney2020learning}{2020})
propose an approach using counterfactuals in training that does not rely
on data augmentation: they argue that counterfactual pairs typically
already exist in training datasets. Specifically, their approach relies
on, firstly, identifying similar input samples with different
annotations and, secondly, ensuring that the gradient of the classifier
aligns with the vector between pairs of counterfactual inputs using the
cosine distance as a loss function (referred to as \emph{gradient
supervision}) (\textbf{\emph{this might be useful for our task as
well}}). In the natural language processing (NLP) domain,
counterfactuals have similarly been used to improve models through data
augmentation: Wu et al. (\citeproc{ref-wu2021polyjuice}{2021}), propose
POLYJUICE, a general-purpose counterfactual generator for language
models. They demonstrate empirically that augmenting training data
through POLYJUICE counterfactuals improves robustness in a number of NLP
tasks.

@@

Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) propose two
definitional differences between Adversarial Examples (AE) and
Counterfactual Explanations (CE): firstly, and more importantly
according to the authors, the term AE implies missclassification, which
is not the case for CE (\textbf{\emph{this might be a useful notion for
use to distinguish between adversarials and explanations during
training}}); secondly, they argue that closeness plays a more critical
role in the context of CE but confess that even counterfactuals that are
not close might be relevant explanations. Pawelczyk et al.
(\citeproc{ref-pawelczyk2022exploring}{2022}) show that CE and AE are
equivalent under certain conditions and derive upper bounds on the
distances between them.

@@

Guo, Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the
first to propose end-to-end training pipeline that includes
counterfactual explanations as part of the training prodeduce. In
particular, they propose a specific network architecture that includes a
predictor and CE generator network (\textbf{\emph{akin a GAN?}}), where
the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network (\textbf{\emph{here we are aligned}}). In
contrast, we impose no restrictions on the neural network architecture
at all. (\textbf{\emph{to ensure the one-hot encoding of categorical
features is maintained, they simple use softmax (might be interesting
for CE.jl)}}) Interestingly, the authors find that their approach is
sensitive to the choice of the loss function: only MSE seems to lead to
good performance. They also demonstrate theoretically, that the
objective function is difficult to optimize due to divergent gradients
and suffers from poor adversarial robustness. (\textbf{\emph{because
partial gradients with respect to the classification loss component and
the counterfactual validity component point in opposite directions}}).
To mitigate these issues, the authors use block-wise gradient descent:
they first update with respect to classification loss and then use a
second update with respect to the other loss components
(\textbf{\emph{this might be useful for our task as well}}). Ross,
Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024}) propose a
way to train models that are guaranteed to provide recourse for
individuals with high probability. The approach builds on adversarial
training (\textbf{\emph{here we are aligned}}), where in this context
adversarial examples are actively encouraged to exist, but only target
attacks with respect to the positive class. The proposed method allows
for imposing a set of actionable recourse ex-ante: for example, users
can impose mutability constraints for features (\textbf{\emph{here we
are aligned}}). (\textbf{\emph{To solve their objective function more
efficiently, they use a first-order Taylor approximation to approximate
the recourse loss component (might be applicable in our case)}})

Luu and Inoue (\citeproc{ref-luu2023counterfactual}{2023}) introduce
Counterfactual Adversarial Training (CAT) with intention of improving
generalization and robustness of language models. Specifically, they
propose to proceed as follows: firstly, identify training samples that
are subject to high predictive uncertainty (entropy); secondly, generate
counterfactual explanations for those samples; and, finally, finetune
the model on the augmented dataset that includes the generated
counterfactuals.

@@

\bibitem[\citeproctext]{ref-goodfellow2014explaining}
Goodfellow, Ian J, Jonathon Shlens, and Christian Szegedy. 2014.
{``Explaining and Harnessing Adversarial Examples.''}
\url{https://arxiv.org/abs/1412.6572}.

\bibitem[\citeproctext]{ref-goodfellow2016deep}
Goodfellow, Ian, Yoshua Bengio, and Aaron Courville. 2016. \emph{Deep
{Learning}}. {MIT Press}.

@@

\bibitem[\citeproctext]{ref-kolter2023keynote}
Kolter, Zico. 2023.{``{Keynote Addresses: SaTML 2023 }.''} In \emph{2023
IEEE Conference on Secure and Trustworthy Machine Learning (SaTML)},
xvi--. Los Alamitos, CA, USA: IEEE Computer Society.
\url{https://doi.org/10.1109/SaTML54575.2023.00009}.

@@

\bibitem[\citeproctext]{ref-rudin2019stop}
Rudin, Cynthia. 2019. {``Stop Explaining Black Box Machine Learning
Models for High Stakes Decisions and Use Interpretable Models
Instead.''} \emph{Nature Machine Intelligence} 1 (5): 206--15.
\url{https://doi.org/10.1038/s42256-019-0048-x}.

@@

\bibitem[\citeproctext]{ref-wilson2020case}
Wilson, Andrew Gordon. 2020. {``The Case for Bayesian Deep Learning.''}
\url{https://arxiv.org/abs/2001.10995}.

### 2025-01-31 paper/sections/abstract.qmd
uff this is going too slow

Counterfactual Explanations (CE) have emerged as a popular method to explain the predictions made by opaque machine learning models in a post-hoc fashion. We propose a novel approach that leverages counterfactuals during the training phase of models.

>>>>

Counterfactual Explanations (CE) have emerged as a popular tool to explain predictions made by opaque machine learning models: they explain how factual inputs need to change in order for some fitted model to produce some desired output. Much existing research has focused on identifying explanations that are not only valid but also deemed desirable with respect to the underlying data and stakeholder requirements. Recent work has shown that under this premise, the task of learning desirable explanations is effectively reassigned from the model itself to the (post-hoc) counterfactual explainer. Building on that work, we propose a novel model objective that leverages counterfactuals during the training phase (ad-hoc) in order to minimize the divergence between learned representations and desirable explanations. Through extensive experiments, we demonstrate that our proposed methodology facilitates training models that inherently deliver desirable explanations while maintaining high predictive performance.
